Fast Fujisaki-Okamoto transformation using encrypt-then-mac and applications to Kyber

### 2024-08-19 etm-transform/manuscript/main.tex
Scaffolding for experimental data

\begin{algorithm}[H]
    \caption{\texttt{Kyber.CCAKEM.KeyGen()}}
    \begin{algorithmic}[1]
        \State $z \leftsample \mathcal{B}^{32}$
            \Comment{Randomly sample 32 bytes (256 bits)}
        \State $(\pk, \sk^\prime) \leftsample \texttt{Kyber.CPAPKE.KeyGen()}$
        \State $\sk = (\sk^\prime, \pk, H(\pk), z)$
            \Comment{H is instantiated with SHA3-256}
        \State \Return $(\pk, \sk)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{\texttt{Kyber.CCAKEM.Encap$^+$(\pk)}}
    \begin{algorithmic}[1]
        \State $m \leftsample \mathcal{B}^{32}$
       % \State $m'= H(m)$
            \Comment{Do not output system RNG directly}
        \State $(\bar{K}, r, K_{mac}) = G(m' \Vert H(\pk))$
            \Comment{G is instantiated with SHA3-512}
        \State $c  \leftarrow \texttt{Kyber.CPAPKE.Enc(\pk, m', r)}$
            \Comment{Because $r$ is set, \texttt{CPAPKE} is deterministic}
            \State $t_1=MAC(K_{mac}, c)$
        \State $K = \texttt{KDF}(\bar{K} \Vert t_1)$
       % \State $t_2=MAC(K, t_1)$
            \Comment{\texttt{KDF} is instantiated with Shake256}
        %\State $(c, t_1, t_2)$
        \State $c \leftarrow (c, t_1)$
        \State \Return $(c, K)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{$\texttt{Kyber.CCAKEM.Decap}^+ (\sk, c, t_1, t_2)$}
    \begin{algorithmic}[1]
        \Require Secret key $\sk = (\sk^\prime, \pk, H(\pk), z)$
        \Require \texttt{Kyber.CPAPKE} Ciphertext $c$ and Tags $t_1, t_2$
        \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
            \Comment{Unpack the secret key; $h$ is the hash of $\pk$}
        \State $(c, t_1) \leftarrow c$
        \State $ m = \texttt{Kyber.CPAPKE.Dec}(\sk^\prime, c)$ 
       % \State $m'=H(m) $ 
        \State $(\overline{K},   r, K_{mac}) =   G(m^\prime \Vert h)$
        \State $t_1'=\mac(K_\mac, c)$
        \If{$t_1' = t_1$}
            \State $K  = \texttt{KDF}(\bar{K} \Vert t_1)$ 
            %\State $t_2'= \mac(K, t_1)$
            \State \Return $K$
         \Else 
            \State Abort
        \EndIf
    \end{algorithmic}
\end{algorithm}

@@

\subsection{Choosing message authentication code}\label{sec:mac-performance}
Given that each keypair is only used once, an adversary can only obtain one authenticated ciphertext, which means that the corresponding MAC key is also only used once per key exchange. This allows us to use MAC constructions that are only one-time secure for substantial computation efficiency gains. We compared the median and mean number of CPU cycles (Intel i7-9750H) used to sign Kyber's IND-CPA ciphertexts. The results are listed in \ref{tbl:standalone-mac-perf}.

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        Name & Security & 768 bytes & 1088 bytes & 1568 bytes \\
        \hline
        CMAC & many-time & 5022 & 5442 & 6090 \\
        & & 5131 & 5578 & 6154 \\
        \hline
        GMAC & one-time & 2778 & 2756 & 2762 \\
        & & 2843 & 2780 & 2919 \\
        \hline
        KMAC-256 & many-time & 7934 & 9862 & 11742 \\
        & & 8594 & 10693 & 12319 \\
        \hline
        Poly1305 & one-time & 1128 & 1218 & 1338 \\
        & & 1435 & 1504 & 1625 \\
        \hline
    \end{tabular}
    \caption{Standalone $\mac$ performances}\label{tbl:standalone-mac-perf}
\end{table}

>>>>

\begin{algorithm}[H]
    \caption{\texttt{Kyber.CCAKEM.KeyGen()}}\label{alg:kyber-ae-keygen}
    \begin{algorithmic}[1]
        \State $z \leftsample \mathcal{B}^{32}$
            \Comment{Randomly sample 32 bytes (256 bits)}
        \State $(\pk, \sk^\prime) \leftsample \texttt{Kyber.CPAPKE.KeyGen()}$
        \State $\sk = (\sk^\prime, \pk, H(\pk), z)$
            \Comment{H is instantiated with SHA3-256}
        \State \Return $(\pk, \sk)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{\texttt{Kyber.CCAKEM.Encap$^+$(\pk)}}\label{alg:kyber-ae-encap}
    \begin{algorithmic}[1]
        \State $m \leftsample \mathcal{B}^{32}$
       % \State $m'= H(m)$
            \Comment{Do not output system RNG directly}
        \State $(\bar{K}, r, K_{mac}) = G(m' \Vert H(\pk))$
            \Comment{G is instantiated with SHA3-512}
        \State $c  \leftarrow \texttt{Kyber.CPAPKE.Enc(\pk, m', r)}$
            \Comment{Because $r$ is set, \texttt{CPAPKE} is deterministic}
            \State $t_1=MAC(K_{mac}, c)$
        \State $K = \texttt{KDF}(\bar{K} \Vert t_1)$
       % \State $t_2=MAC(K, t_1)$
            \Comment{\texttt{KDF} is instantiated with Shake256}
        %\State $(c, t_1, t_2)$
        \State $c \leftarrow (c, t_1)$
        \State \Return $(c, K)$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{$\texttt{Kyber.CCAKEM.Decap}^+ (\sk, c, t_1, t_2)$}\label{alg:kyber-ae-decap}
    \begin{algorithmic}[1]
        \Require Secret key $\sk = (\sk^\prime, \pk, H(\pk), z)$
        \Require \texttt{Kyber.CPAPKE} Ciphertext $c$ and Tags $t_1, t_2$
        \State $(\sk^\prime, \pk, h, z) \leftarrow \sk$
            \Comment{Unpack the secret key; $h$ is the hash of $\pk$}
        \State $(c, t_1) \leftarrow c$
        \State $ m = \texttt{Kyber.CPAPKE.Dec}(\sk^\prime, c)$ 
       % \State $m'=H(m) $ 
        \State $(\overline{K},   r, K_{mac}) =   G(m^\prime \Vert h)$
        \State $t_1'=\mac(K_\mac, c)$
        \If{$t_1' = t_1$}
            \State $K  = \texttt{KDF}(\bar{K} \Vert t_1)$ 
            %\State $t_2'= \mac(K, t_1)$
            \State \Return $K$
         \Else 
            \State Abort
        \EndIf
    \end{algorithmic}
\end{algorithm}

@@

\subsection{MAC performance}\label{sec:mac-performance}
When instantiating an instance $\kem(\keygen, \encap, \decap)$, there are a variety of possible MAC's to choose from. For each of the security level and choice of MAC, we measured the number of CPU cycles needed to produce a digest of the unauthenticated ciphertext. The median (top) and mean (bottom) measurements are reported in table \ref{tbl:standalone-mac-perf}.

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Name & Security & measurement & 768 bytes & 1088 bytes & 1568 bytes \\
        \hline
        CMAC & many-time & median & 5022 & 5442 & 6090 \\
        & & mean & 5131 & 5578 & 6154 \\
        \hline
        KMAC-256 & many-time & median & 7934 & 9862 & 11742 \\
        & & mean & 8594 & 10693 & 12319 \\
        \hline
        GMAC & one-time & median & 2778 & 2756 & 2762 \\
        & & mean & 2843 & 2780 & 2919 \\
        \hline
        Poly1305 & one-time & median & 1128 & 1218 & 1338 \\
        & & mean & 1435 & 1504 & 1625 \\
        \hline
    \end{tabular}
    \caption{Standalone $\mac$ performances}\label{tbl:standalone-mac-perf}
\end{table}

Based on the security reduction in section \ref{sec:main-results} we chose standalone GMAC and Poly1305 for their substantial performance advantage over other constructions such as CBC-MAC and KMAC. We then modified the reference implementation (TODO: citation needed https://github.com/pq-crystals/kyber) according to algorithms \ref{alg:kyber-ae-keygen}, \ref{alg:kyber-ae-encap}, and \ref{alg:kyber-ae-decap}. We measured the median and mean CPU cycles needed to run each of the routines. The measurements are listed in table \ref{tbl:kyber-ae-perf}

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Name & Security level & measurement & $\keygen$ & $\encap$ & $\decap$ \\
        \hline
        Kyber512 & 128 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        Kyber768 & 192 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        Kyber1024 & 256 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE512 w/ GMAC& 128 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE768 w/ GMAC& 192 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE1024 w/ GMAC& 256 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE512 w/ Poly1305& 128 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE768 w/ Poly1305& 192 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
        KyberAE1024 w/ Poly1305& 256 bits & median &  & &  \\
        & & mean &  & &  \\
        \hline
    \end{tabular}
    \caption{Kyber-AE performance measurements}\label{tbl:kyber-ae-perf}
\end{table}
